---
title: "Cyclistic Analysis"
author: "Gentian Gashi"
date: "2022-08-12"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/gashi/Desktop/Google Data Analytics Case Study (Cyclistic)/Raw Data")
```

# Cyclistic Business Case Analysis

In this scenario I am a junior data analyst at the fictional company Cyclistic and I have been tasked with the overall goal of designing marketing strategies for Cyclistic, a bike sharing company. Cyclistic has provided me with a large dataset of bike rides that have happened over the past 12 months. With this analysis, I will show the different types rides that are recorded by Cyclistic, who rides the bikes, and more descriptive statistics.

This file is intended to only showcase the R code used to clean and analyze the data. For an in-depth analysis, check out my Medium page, referenced in the README file on Git.

***

## Data Processing and Cleaning
### Setting Work Directory and Loading Packages

```{r}
setwd("C:/Users/gashi/Desktop/Google Data Analytics Case Study (Cyclistic)/Raw Data")
library(tidyverse) #transform data
library(lubridate) #work with dates
library(hms) #work with hours
library(ggplot2) #graphs
library(scales) #scales for graphs
library(gtExtras) #fancy tables
library(gt) #fancy tables
library(webshot) #exporting fancy tables to jpg
#webshot::install_phantomjs()
```

### Importing Data

Since there are several datasets, instead of importing them manually, I will do so with a for loop. However, I will import them as individual files because the names of the columns might differ between files and that would present a problem when trying to import them in one file.

```{r}
#Read files named xyz1111.csv, xyz2222.csv, etc. in working directory
filenames <- list.files(pattern="*csv")

##Create list of data frame names without the ".csv" part 
names <-substr(filenames,1,21)

###Load all files
for(i in names){
    filepath <- file.path("C:/Users/gashi/Desktop/Google Data Analytics Case Study (Cyclistic)/Raw Data",paste(i,".csv",sep=""))
    assign(i, read.csv(filepath, sep = ",")) #the data is read in via read.csv and to change the column types, add col_type for each variable
}
```

### Checking the column names to see if they match and allow appending the data
```{r}
colnames(`202108-divvy-tripdata`)
colnames(`202109-divvy-tripdata`)
colnames(`202110-divvy-tripdata`)
colnames(`202111-divvy-tripdata`)
colnames(`202112-divvy-tripdata`)
colnames(`202201-divvy-tripdata`)
colnames(`202202-divvy-tripdata`)
colnames(`202203-divvy-tripdata`)
colnames(`202204-divvy-tripdata`)
colnames(`202205-divvy-tripdata`)
colnames(`202206-divvy-tripdata`)
colnames(`202207-divvy-tripdata`)
```

### Appending the Data
```{r}
# Appending the data
cyclist <- rbind(`202108-divvy-tripdata`, `202109-divvy-tripdata`, `202110-divvy-tripdata`, `202111-divvy-tripdata`,
                 `202112-divvy-tripdata`, `202201-divvy-tripdata`, `202202-divvy-tripdata`, `202203-divvy-tripdata`,
                 `202204-divvy-tripdata`, `202205-divvy-tripdata`, `202206-divvy-tripdata`, `202207-divvy-tripdata`)

str(cyclist) # seeing the column types
```

### Cleaning and Inspecting the Data
```{r}
#Removing unnecessary columns
df2 <- cyclist %>% 
    select(-start_lat, -start_lng, -end_lat, -end_lng, -start_station_name, -start_station_id, -end_station_name, -end_station_id)


colnames(df2)
nrow(df2)
dim(df2)
head(df2)
str(df2)
summary(df2)
df2 %>% count(member_casual)
df2 %>% count(rideable_type)
```

### Generating Necessary Variables
```{r}
df3 <- df2 %>% 
    mutate(year = year(started_at),
           month = month(started_at, label = T),
           day = day(started_at),
           day_of_week = wday(started_at, week_start = getOption("lubridate.week.start", 1), label = T), #starting the week from Monday (default Sunday - 7)
           date = date(started_at),
           ride_length = round(difftime(ymd_hms(ended_at), ymd_hms(started_at), units = "mins"), digits = 1))
```

### Final Cleaning of the Data
```{r}
df3 <- na.omit(df3) #remove rows with NA values
df3 <- distinct(df3) #remove duplicate rows 
df3 <- df3[!(df3$ride_length <=0),] #remove where ride_length is 0 or negative
```

***

## Descriptive Analysis
### Analysis
```{r}
#Number of rides per type of member
df3 %>% 
    group_by(member_casual) %>% 
    summarize(count=n()) %>% 
    mutate(percent = count/sum(count)*100)

#Summary statistics per type of member
df3 %>% 
    group_by(member_casual) %>% 
    summarize(mean = mean(ride_length),
              median = median(ride_length),
              min = min(ride_length),
              max = max(ride_length))

#Average ride time by each day for members vs casual users
df3 %>% 
    group_by(member_casual, day_of_week) %>% 
    summarize(mean = mean(ride_length))

#Average ride time by each day for members vs casual users
df3 %>% 
    group_by(member_casual, day_of_week) %>% 
    summarize(number_of_rides=n(),
              mean_ride_length = mean(ride_length))
```

### Visualisations
#### General Infosheet
```{r}
#Infosheet Table
library(data.table) #creating data tables
atb <- c("Most Rides by Type", "Average Ride Length", "Busiest Month", "Busiest Weekday", "Most Used Bike")
info <- c("Member", "20 Minutes", "July", "Saturday", "Classic" )
infosheet <- data.table(atb, info)


infosheet %>% 
    gt() %>% 
    gt_theme_guardian() %>% 
    tab_style(locations = cells_body( #changes cells
        rows = everything(),
        columns = "info"
    ),
    style = list(cell_text(color = "#7d3ac1",
                           size = "medium",
                           weight = "bold"))
    ) %>% 
    tab_options(column_labels.hidden = TRUE)
```

#### Pie Chart of Total Rides
```{r}
#Pie chart of Total Rides
df3 %>% 
    group_by(member_casual) %>% 
    summarize(number_of_rides = n()) %>% 
    ggplot(aes(x="", y=number_of_rides, fill=member_casual)) +
    geom_col(alpha=0.7, width=0.3) +
    coord_polar(theta = "y") + #makes the pie chart
    scale_fill_manual(values=c("#ea7369","#7d3ac1"), breaks = c("casual", "member"), labels=c("Casual", "Member")) +
    ggtitle("Total Rides by Subscription") +
    annotate("text", x = 0, y = 0, label = "5.9M Total Rides", colour="black", fontface="bold", size=7) + #adds text in the middle
    theme_minimal() +
    theme(legend.title = element_blank(),
          legend.text = element_text(face="bold"),
          axis.title = element_blank(),
          axis.text = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.margin = margin(1,1,1.5,1.2, "cm"),
          plot.title = element_text(face="bold"))
```

#### Total Rides Table
```{r}
#Total Rides Table
df3 %>% 
    group_by(member_casual) %>% 
    summarize(count=n()) %>% 
    mutate(percent = paste0(round((count/sum(count)*100), 1), "%"),
           count2 = label_comma()(count),
           member_casual = case_when(member_casual == "casual" ~ "Casual",
                                     member_casual == "member" ~ "Member")) %>%
    select(-count) %>%
    relocate(count2, .after=member_casual) %>% 
    rename(Subscriber = member_casual,
           'Total Rides' = count2,
           'Percentage of Total' = percent) %>% 
    gt() %>% 
    gt_theme_guardian() %>% 
    data_color(
        columns = 'Percentage of Total', colors = c("#7d3ac1"))
```

#### Average Ride Length Table
```{r}
#Average Ride Length Table
df3 %>% 
    group_by(member_casual) %>% 
    summarize(mean=mean(ride_length)) %>% 
    mutate(
        avg = paste0(                    #paste0 pastes better than just paste
            round(mean, 1)," ","Minutes"
        ),
        member_casual = case_when(member_casual == "casual" ~ "Casual",
                                  member_casual == "member" ~ "Member")
    ) %>%
    select(-mean) %>% 
    rename(Subscriber = member_casual,
           'Avg. Ride Length' = avg) %>% 
    gt() %>% 
    gt_theme_guardian() %>% 
    data_color(
        columns = 'Avg. Ride Length', colors = c("#7d3ac1"))
```


#### Bar Graph of Total Rides per Weekday
```{r}
#Bar Chart of total rides by day of week (value labels are shown fully)    
df3 %>% 
    group_by(member_casual, day_of_week) %>% 
    summarize(number_of_rides = n()) %>% 
    ggplot(aes(x=day_of_week, y=number_of_rides, fill=member_casual)) +
    geom_chicklet(position = "stack", alpha=0.7, radius = grid::unit(5, "mm"), width = 0.6) +
    scale_y_continuous(labels = unit_format(unit = "K", scale = 1e-3)) +
    #    scale_fill_discrete(labels=c("Casual", "Member")) +
    scale_fill_manual(values=c("#ea7369","#7d3ac1"), breaks = c("casual", "member"), labels=c("Casual", "Member")) +
    geom_text(aes(label = label_comma()(number_of_rides)), position = position_stack(vjust = 1.05), colour="black", fontface = "bold") + #label values shown fully and with a comma
    ggtitle("Total Rides by Subscription and Day of Week") +
    labs(x = "Day of the week",
         y = "Number of rides",
         fill = "Type of Member") +
    theme_minimal() +
    theme(legend.title = element_blank(),
          legend.text = element_text(face="bold"),
          axis.title = element_blank(),
          axis.text = element_text(face="bold"),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.margin = margin(1,1,1.5,1.2, "cm"),
          plot.title = element_text(face="bold"))
```

#### Bar Graph of Total Rides per Bike Type
```{r}
#Bar Chart of total rides by bike type (value labels are shown fully)    
df3 %>% 
    group_by(member_casual, rideable_type) %>% 
    summarize(number_of_rides = n()) %>% 
    ggplot(aes(x=rideable_type, y=number_of_rides, fill=member_casual)) +
    geom_chicklet(position = "stack", alpha=0.7, radius = grid::unit(5, "mm"), width = 0.6) +
    coord_flip() +
    scale_y_continuous(labels = unit_format(unit = "K", scale = 1e-3)) +
    scale_fill_manual(values=c("#ea7369","#7d3ac1"), breaks = c("casual", "member"), labels=c("Casual", "Member")) +
    scale_x_discrete(labels = c("Classic Bike", "Docked Bike", "Electric Bike")) + #labels axis ticks manually 
    geom_text(aes(label = label_comma()(number_of_rides)), position = position_stack(vjust = 0.5), colour="black", fontface="bold") + #label values shown fully
    ggtitle("Total Rides by Subscription and Bike Type") +
    labs(x = "Bike Type",
         y = "Number of rides",
         fill = "Type of Member") +
    theme_minimal() +
    theme(legend.title = element_blank(),
          legend.text = element_text(face="bold"),
          axis.title = element_blank(),
          axis.text = element_text(face="bold"),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.margin = margin(1,1,1.5,1.2, "cm"),
          plot.title = element_text(face="bold"))
```

#### Bar Graph of Total Rides per Month
```{r}
#Bar Chart of total rides by month (value labels are shown fully)    
df3 %>% 
    group_by(member_casual, month) %>% 
    summarize(number_of_rides = n()) %>% 
    ggplot(aes(x=month, y=number_of_rides, fill=member_casual)) +
    geom_chicklet(position = "stack", alpha=0.7, radius = grid::unit(3, "mm"), width = 0.6) +
    scale_y_continuous(labels = unit_format(unit = "K", scale = 1e-3)) + #transforms axis label in thousands
    scale_fill_manual(values=c("#ea7369","#7d3ac1"), breaks = c("casual", "member"), labels=c("Casual", "Member")) +
    geom_text(aes(label = label_comma()(number_of_rides)), position = position_stack(vjust = 0.5), colour="black", fontface="bold") + #label values shown fully
    ggtitle("Total Rides by Month") +
    labs(x = "Month",
         y = "Number of rides",
         fill = "Type of Member") +
    theme_minimal() +
    theme(legend.title = element_blank(),
          legend.text = element_text(face="bold"),
          axis.title = element_blank(),
          axis.text = element_text(face="bold"),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.margin = margin(1,1,1.5,1.2, "cm"),
          plot.title = element_text(face="bold"))
```

